---
title: "Variability in tree height and diameter across different
species in Alaska, USA"
author: "Vignesh Arunkumar, Aidan Power, Brynn Rotbart, Sophie Valkenberg"
date: "2024-12-03"
output: html_document
---
```{r setup, include=FALSE}
#Loading Packages
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse) 

```


```{r}
# Set your working directory
getwd()
setwd("/home/guest/FinalEDEProject/EDEProj_Arunkumar_Power_Rotbart_Valkenberg")

```


```{r}
# Load your datasets
Ak_data <- 
  read.csv('~/FinalEDEProject/EDEProj_Arunkumar_Power_Rotbart_Valkenberg/AK_SITETREE.csv')
```

# Rationale and Research Questions
How does tree height vary with diameter across different species in Alaska's
forests?



```{r}
#Filtering for Species
Species.wanted <- Ak_data %>%
  filter(SPCD %in% c(94, 95, 108, 264))
Species.wanted <- Species.wanted %>%
  mutate(Species = case_when(
    SPCD == 94 ~ "White Spruce",
    SPCD == 95 ~ "Black Spruce",
    SPCD == 108 ~ "Lodgepole Pine",
    SPCD == 264 ~ "Mountain Hemlock"
  ))
```


```{r}
# Set your ggplot theme


```


```{r}
#Linear Regression to see correlation between HT and DIA for all Species in Data
All.species.AK <- lm(HT ~ DIA, data = Ak_data)
summary(All.species.AK)

ggplot(Ak_data, aes(x = DIA, y = HT)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Tree Height vs Diameter in Alaska",
       x = "Diameter",
       y = "Height")

```





```{r}
#Linear Regression to see correlation between HT and DIA taking into account 
#four species. 
species.model <- lm(HT ~ DIA + Species, data = Species.wanted)
summary(species.model)

ggplot(Species.wanted, aes(x = DIA, y = HT, color = factor(Species))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Tree Height vs Diameter by Species",
       x = "Diameter",
       y = "Height",
       color = "Species Code")

ggplot(Species.wanted, aes(x = Species, y = HT, color = factor(Species))) +
  geom_boxplot()+
  labs(title = "Tree Height by Species",
       x = "Species",
       y = "Height",
       color = "Species Code")

```

(The other 3 species being compared to black spruce: This can be deleted just letting you know)




```{r}
#Mean Height of Black Spruce compared to Mean Height of White Spruce
b.w.spruces <- Species.wanted %>%
  filter(SPCD %in% c(94, 95))
BlackvWhiteSpruce <- aov(HT~Species, data = b.w.spruces)

summary.aov(BlackvWhiteSpruce)
```




\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage






\newpage

# Dataset Information



\newpage

# Exploratory Analysis 



\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 








