---
title: "Variability in tree height and diameter across different
author: "Vignesh Arunkumar, Aidan Power, Brynn Rotbart, Sophie Valkenberg"
date: "2024-12-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Setting up our R Workspace
```{r setup, include=FALSE}
#Loading Packages
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse)

```


```{r}
# Set your working directory
getwd()
setwd("/home/guest/FinalEDEProject/EDEProj_Arunkumar_Power_Rotbart_Valkenberg")

```


```{r}
# Load your datasets
Ak_data <- 
  read.csv('~/FinalEDEProject/EDEProj_Arunkumar_Power_Rotbart_Valkenberg/AK_SITETREE.csv')
```

# Rationale and Research Questions
  We all chose the topic of tree heights because it is one of the main things that our foursome has in common, as two of us are in the forestry program, while the other two are TFE. In addition, we all found a similar interest of not only how tree heights vary with tree diameter and species, but also these factor in trees of a region unknown to us. 


	The reason we chose this particular dataset is because of none of us have any experience traveling or even reading about Alaska. Our ignorance, combined with the dataset including height, DBH, and species codes, motivated us to use this dataset to answer our questions. Additionally, the data comes from the United States Forest Service, which is a group we all have learned much about and wanted to interact with. 

Q: How does tree height vary across Alaska’s forests?
-	How does this height vary with DBH?
-	How does this height vary with species?


# Data Wrangling 
```{r}
#Filtering for Species
Species.wanted <- Ak_data %>%
  filter(SPCD %in% c(94, 95, 108, 264))
Species.wanted <- Species.wanted %>%
  mutate(Species = case_when(
    SPCD == 94 ~ "White Spruce",
    SPCD == 95 ~ "Black Spruce",
    SPCD == 108 ~ "Lodgepole Pine",
    SPCD == 264 ~ "Mountain Hemlock" 
  ))
Species.wanted$Species <- as.factor(Species.wanted$Species)
Species.wanted$PLOT <- as.factor(Species.wanted$PLOT)
```

## GGPlot Theme Setup 
```{r}
# Set your ggplot theme


```


# Linear Regressions
## Correlation between Height and Diameter for all Species
```{r}
#Linear Regression to see correlation between HT and DIA for all Species in Data
All.species.AK <- lm(HT ~ DIA, data = Ak_data)
summary(All.species.AK)

ggplot(Ak_data, aes(x = DIA, y = HT)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Tree Height vs Diameter in Alaska",
       x = "Diameter",
       y = "Height")

```




## Correlation between Height and Diameter taking Species of Interest into Account
```{r}
#Linear Regression to see correlation between HT and DIA taking into account 
#four species. 
species.model <- lm(HT ~ DIA + Species, data = Species.wanted)
summary(species.model)

ggplot(Species.wanted, aes(x = DIA, y = HT, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Tree Height vs Diameter by Species",
       x = "Diameter",
       y = "Height",
       color = "Species Code")

```

## Correlation between Height and Diameter taking Species of Interest and Plot ID into Account
```{r}
#Linear regression to see correlation between HT and DIA taking into account
#four species and plot ID
PlotandSpecies.model <- lm(HT ~ DIA + Species + PLOT, data = Species.wanted)
summary(PlotandSpecies.model)

#can't visualize this linear regression because there's too many different #s of plots (can delete after y'all read this)

```

## ANOVA Tests
```{r}
#ANOVA tests for HT and DIA correlation

#ANOVA with species interaction
species.model.anova <- aov(HT ~ DIA + Species, 
                           data = Species.wanted)
summary(species.model.anova)

#ANOVA with species and plot interaction
PlotandSpecies.model.anova <- aov(HT ~ DIA + Species + PLOT, 
                                  data = Species.wanted)
summary(PlotandSpecies.model.anova)

```

The ANOVA results suggest that diameter has the highest influence on tree height, as expected. Additionally, it suggests that both species and plot number have a significant influence on tree height. The influence of plot is smaller than that of diameter and tree species. These results will be explored further in the following sections. 


## Mean Height Comparisons 
```{r}
#Boxplot of Tree Height by Species
ggplot(Species.wanted, aes(x = Species, y = HT, color = Species)) +
  geom_boxplot()+
  labs(title = "Tree Height by Species",
       x = "Species",
       y = "Height",
       color = "Species Code")

#(The other 3 species being compared to black spruce: This can be deleted just letting you know)

#Mean Height of Black Spruce compared to Mean Height of White Spruce
b.w.spruces <- Species.wanted %>%
  filter(SPCD %in% c(94, 95))
BlackvWhiteSpruce <- aov(HT~Species, data = b.w.spruces)

summary.aov(BlackvWhiteSpruce)

```

## Visualizing Interactions with Plot Differences
```{r}
random_plot_subset <- Species.wanted[sample(nrow(Species.wanted), size = 15, 
                                            replace = FALSE), ]

#Plot showing relationship between PLOT and HT (random sample)
ggplot(random_plot_subset, aes(x = PLOT, y = HT)) +
  geom_boxplot(aes(color = PLOT)) +
  labs(title = "Tree Height by Plot (Random Sample)",
       x = "Plot",
       y = "Height") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(legend.position = "none")

#Plot showing relationship between HT, PLOT, and Species (random sample)
ggplot(random_plot_subset, aes(x = Species, y = HT, color = PLOT)) +
  geom_boxplot() +
  labs(title = "Tree Height by Species and Plot (Random Sample)",
       x = "Species",
       y = "Height",
       color = "Plot Code") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "bottom")

```



\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage






\newpage

# Dataset Information
Data was collected...

Once the data was added to our catalog we...




\newpage

# Exploratory Analysis 



\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 








